### 2011-10-13 doc/manuscript.tex
Just an update.

\section*{Abstract}
Homologous recombination and horizontal gene transfer are investigated with
complete genomes of closely related bacterial species of \textit{Streptococcus
pyogenes} (SPY) and \textit{S.\ dysgalactiae} subspecies \textit{equisimilis}
(SDE).  The former is a pervasive human-specific pathogen, and the latter an
opportunistic human pathogen.  We aims at which parts of the genomic
organization and human pathogenic virulence factors of SDE could have originated
in those of SPY, or vice versa.  Our genome-wide survey of gene transfer
indicated that gene transfer via homologous recombination occurred more for the
net direction from SPY to SDE.  Although we did not find enrichment of virulence
genes in homologous recombining gene transfer, we found evidence that
virulence genes transferred more often via non-recombining fashion without
replacing parts of destination genomes than did non-virulence genes.  A
genome-wide resource of recombining and non-recombining gene transfer for the
study of the SPY and SDE genomes is provided so that research community in
pathogenic bacteria could take advantage of it.

@@

Bacteria evolves new biological features through horizontally transferred genes
from other lineages and vertically transmitted mutations on genes within the
same lineage.  Mutations on a gene modifiy relatively small portions of the gene
through substitutions, insertions, and deletions of deoxyribonucleic acids.
Gene transfers can reshape the organization of a bacterial genome,
which could revamp biological functions of the species.  Recipient bacteria of
transferred genes can substitute transferred genes for homologous parts of their
genomes, or integrate transferred genes into their genomes without replacement.
These two processes are here in this manuscript discriminated by calling the
former \textit{recombining gene transfer}, and the latter \textit{horizontal
gene transfer} \citep{Ochman2001,Lawrence2009}.  Despite the two distinct
evolutionary processes in bacteria, analyses of complete genome sequences in
efforts of better understanding bacterial evolution have paid less attention to
a combined approach to illuminating disparity of the two gene transfer
processes. We attempted here to use and compare two approaches for
distinguishing recombining gene transfer process from horizontal, or
non-recombining, gene transfer, and tackled their roles in association with
virulence genes.

@@

In short, the complete genomes of the two spcies of SPY and SDE presented us a
unique opportunity to study gene transfer over relatively recent time scales.
We took advantage of the model-based approach for inferring homologous
recombination, and the parsimony-based one for detecting gene duplication, loss,
and horizontal transfer. We compared both of the results in recombining gene
transfer events to show that the parimony-based method inferred recombining gene
transfer events in biological functional categories similar to those inferred
with the model-based approach. This was the first attempt as far as we know to
analyze SPY and SDE in genome-wide scale for detecting gene transfer events not
only in the core genome but also in the pan genome, which makes the analysis
complete in terms of the amount of used genome data. Our analysis of core
genomes also showed for the first time that the net directionality of
recombining gene transfer from SPY to SDE was stronger than the reverse
direction.

@@

@@

We present a short summary of the statistical method developed by
\citet{Didelot2007} and \citet{Didelot2010} for studying bacterial
recombination. A detailed account was found in the two papers and in the user
guide available at \url{http://code.google.com/p/clonalorigin}.  We used
\texttt{progressiveMauve} v2.3.1 \citep{Darling2004,Darling2010} with default
options to attempt to align the six genomes, which led to too small proportion
of a core genome. We had to remove the SEE sample from the genome alignment
although we kept it in the parsimony-based approach.  We then used
\texttt{stripSubsetLCBs}, a program of the \texttt{ClonalOrigin} package, and
identified 276 blocks of sequence alignments with length threshold of 1500 base
pairs.  We excluded two of these blocks in a subsequent analysis because they
contained long consecutive gaps.  We applied \texttt{ClonalFrame} v1.1 available
at \url{http://www.xavierdidelot.xtreemhost.com} to the remaining 274 blocks,
and estimated the species tree relating the five genomes. \texttt{ClonalFrame}
estimates the average recombinant tract length and the ratio between
recombination and mutation rates, which were not used in the following analysis.
We used a single Markov chain Monte Carlo (MCMC) with 10000 burn-in generations
and 10000 sampling generations to obtain the tree estimate with branch lengths,
and additionally seven independent MCMC runs to validate adequate convergence of
the first run.

@@

\subsection{Local gene trees and recombination intensity}
Because the procedure of summarizing posterior sample of \textit{recombinant
tree} was not detailed in the user guide at the time when we used
\texttt{ClonalOrigin}, we elected to use Figure S\ref{fig:clonalorigin} to
illustrate a recombinant tree in order to clarify summarization of
\textit{recombinant history} posterior samples.  A recombination history is
represented by a \textit{recombinant tree}, which is a species tree augmented
with \textit{recombinant edges}.  Each recombinant edge is defined by the time
and branch of origin and destination, and the genomic interval that it affects.
The branch of origin is referred to as the \textit{donor} and the destination
branch is referred to as the \textit{recipient}. The type of a recombinant edge
is determined by the donor and recipient species tree branches.  A recombinant
tree can be viewed as a restricted version of an ancestral recombination graph
(ARG), and as such it induces a local genealogy for each site in the genomic
block it describes.  \citet{Didelot2010} presented two procedures of summarizing
recombinant trees by viewing trees from two different angles: by counting types
of recombinant edges regardless of genomic positions affected by the edges, and
by measuring recombinant edge changes to the species tree in site-wise manner.
By similarly following their approaches we quantified additional aspects of
recombinant trees; we counted gene tree topologies for each site, and also
quantified recombination intensity along the genome.

The recombination intensity allowed us to find genomic regions enriched for
recombination events in general, or regions enriched for recombination events of
certain types (e.g., between SPY and SDE).  We note that the recombination
intensity is not normalized to the interval $[0,1]$. Rather than giving
probabilistic meanings to the recombination intensity we interpreted it as an
average number of recombinant edge types per site.  The measure appeared to act
as an effective indicator of recombination intensity (see subsection of
\textbf{Simulation Study}).  Our simulation study indicated that edges of
certain types were better indicators of true recombination events than others.
For instance, recombinant edges which resulted in a recombinant tree with the
same topology as the species tree appeared to be more prone to spurious
sampling.  Such edges were ones where the donor was the same branch, sister
branch, or parent branch of the recipient branch.  Therefore, in some cases we
elected to remove such edges from consideration when measuring recombination
intensity.  The tree topologies changes would be correlated with recombination
intensity.  The tree topologies would show more effectively regions where
recombination occurs along the genome (see subsection of \textbf{UCSC Genome
Browser}).

@@

While \texttt{ClonalOrigin} could be used to understand recombining gene
transfers within the conserved core genome, we sought to understand the history
of the remainder of the genome where more complex evolutionary processes occur,
such as gene duplications, losses, and horizontal transfers.  In recent years,
several methods have been developed for inferring these evolutionary events in
gene families by reconciling gene trees with a species tree
\citep{David2011,Doyon2011,Tofigh2011}.  In this analysis, we complemented our
study of recombination by using \texttt{Mowgli} \citep{Doyon2011} to
characterize events that can change the copy number of a gene, such as
duplications, losses, and non-recombining horizontal gene transfers.  In
contrast to \texttt{ClonalOrigin}, \texttt{Mowgli} assumes a parsimony model for
inferring events and uses pre-determined gene clusters where genes are treated
as atomic units.  The advantage of using \texttt{Mowgli} was that it did not
assume the sequence data contains a single copy of each homologous segment, and
it attempted to explain different copy numbers by a series of gain and loss
events.  We here were interested in the approach of \texttt{Mowgli} to study
non-recombining horizontal gene transfers.  We obtained gene family clusters
defined in \citet{Suzuki2011} using the OrthoMCL method \citep{Li2003} over the
full set of six genomes including \textit{S.\ equi} subsp.\textit{\ equi} strain
4047 (SEE).  For each family cluster, the corresponding protein sequences were
aligned using the MUSCLE program \citep{Edgar2004a} and nucleotide alignments
were constructed by mapping nucleotide sequences onto each protein sequence
alignment. We then identified recombination within gene alignments using the
Single Breakpoint Recombination (SBP) method from the HyPhy package
\citep{KosakovskyPond2006}.  Using the Akaike Information Criterion (AIC), SBP
inferred that 47.5\% (1,094) of gene families contained at least one
topology-changing recombination. Each of these families was then split into two
gene fragments.  Gene trees were constructed from nucleotide alignments for each
gene fragment using the RAxML program \citep{Stamatakis2006}, and were rooted by
minimizing inferred duplications and losses.  Each rooted gene tree was then
reconciled using the \texttt{Mowgli} program \citep{Doyon2011}, which inferred
the most parsimonious gene duplication, loss, and transfer events for each
family.  In the case of multiple maximum parsimonious reconciliations, one was
chosen uniformly at random.  Unless specified otherwise, default setings were
used for all programs.  Since the method of \texttt{Mowgli} did not directly
model recombination events, recombination-based transfers were inferred in a
post-processing step, by checking whether the transfer was coupled with a loss
of a homologous gene in the recipient branch (indicating gene replacement rather
than gain).  A transfer event was classified as non-recombining if the
destination species of that transfer had an extant descendant that contained
copies of that genes which were descendants of the transferred gene as well as
copies which were non-descendants.  We noted that this parsimonious approach was
imperfect and would tend to undercount events, especially recombining gene
transfer events, which were modeled using two events: a transfer and a loss.
Nevertheless, this analysis allowed us to study non-recombining gene transfers,
which were not modeled by \texttt{ClonalOrigin}, and to validate some of the
\texttt{ClonalOrigin} results on recombining gene transfer.

@@

% FIXME
% \texttt{The recombination intensity scores were taken to be average site-wise
% intensity scores. SC: THIS NEEDS TO BE REANALYZED WITH RECOMBINATION INTENSITY
% WITH ONLY TREE-TOPOLOGY CHANGING RECOMBINANT EDGES.} 
We performed a series of analyses to check for functional gene categories
enriched for certain types of gene transfer events. Functional categories were
assigned in the same manner as \citet{Suzuki2011}. In short, blastP was used to
compare Streptococcus genes to bacterial proteins from the Uniref90 database.
Matches with E-values less than $1.0\times10^{-5}$ were assumed to have the same
Gene Ontology (GO) classification as the target gene, as determined from the
uniProt GOA database.  We also classified a category of  \textit{virulent genes}
using the virulence profile described in \citet{Suzuki2011}. Among 207 virulent
genes at least one of the genes in the gene family was identified by that
profile.  Each gene covered by our analysis was associated with a recombination
intensity score from the \texttt{ClonalOrigin} analysis and scores from the
\texttt{Mowgli} analysis corresponding to number of duplications, losses,
horizontal gene transfers, and recombining gene transfer events.  Mann-Whitney
tests were used to identify GO categories whose genes showed significantly
elevated values in each score.  A 5\% False Discovery Rate (FDR) correction
\citep{Benjamini1995} was used to determine an appropriate significance
threshold for our p-values.

@@

We conducted experiments on simulation data to check the accuracy of estimates
obtained by \texttt{ClonalOrigin} in a scenario similar to ours. We used a
simulation feature of \texttt{ClonalOrigin} to simulate the process of mutation
and recombination across the five species.  While populaiton mutation and
recombination parameters were accurately estimated, the recombinant tract length
was over-estimated  (Table S\ref{tab:sim-three-population-parameters}).  This
was not surprising as recombinant tract lengths were known to be difficult to
estimate from relatively short alignment blocks \citep{Didelot2010}. 
% When we
% performed simulations which much longer blocks (10000 bp), \texttt{ClonalOrigin}
% was able to estimate the average recombinant tract length accurately (data not
% shown).  
% FIXME: Show this result of 10,000 blocks.
We also evaluated the accuracy of the statistics estimated in the second stage
of \texttt{ClonalOrigin}. The number of recombinant edges for each pair of
recombinant tree branches was correlated with the true value in general while
there were slight over-estimation (Figure S\ref{fig:h1}).  However, this
comparison between inferrence from simulated data and that of real data set
suggested that the directional difference between SPY and SDE was due to neither
lack of power of \texttt{ClonalOrigin} nor hidden bias thereof.  Recombination
intensities that were measured on genes also showed rough correlation (Pearson's
correlation coefficient of 0.63) of the estimated and true values (Figure
S\ref{fig:ri1}).  Estimated values appeared to be driven somewhat by the prior
implied by the population parameter setting, explaining why lower values were
over-estimated and higher values were under-estimated.  Nonetheless, the
correlation indicated that the measure of recombination intensity used in our
functional gene category served as an adequate indicator of actual intensity of
recombination events along genes.

@@

@@

% FIXME: Newick tree in text in a file.
We describe the detailed procedure of simulations for checking if the three
population parameters were accurately recovered from the first stage of
\texttt{ClonalOrigin} MCMC.  Ten recombinant trees were sampled given the three
parameters and the species tree  set to the values inferred from the five
bacterial genomes (see section \textbf{Results}).  Each of the recombinant trees
was used to simulate the Jukes-Cantor model for DNA sequence evolution to create
274 blocks of alignments.  The block lengths configuration was the same as that
of the blocks used in the analysis of the five genomes.  The mean of the ten
estimates of the recombinant tract length was significantly higher than the true
value of (919 bp versus 744 bp) whereas the values of the mutation and
recombination rates were accurately estimated (Table
S\ref{tab:sim-three-population-parameters}).  Recombinant tract lengths were
known to be difficult to estimate from relatively short alignment blocks
\citep{Didelot2010}. Performing similar simulations with much longer blocks
(i.e., 10000 bp) we estimated the average recombinant tract length accurately
\texttt{(Table S data not shown)}.

@@

% FIXME 1: IG: RESULTS FROM THIS POINT ON REQUIRE SOME REFINEMENT. SC: Let's see.
% FIXME 2: ``TRUE" TO BE EXPECTED VALUE
% RATHER THAN AVERAGE OVER THE 10 REC.  TREES. SC: OKAY. LET'S FIND THE PRIOR
% EXPECTED NUMBER OF EVENTS. THIS MUST BE GIVEN SOMEWHERE}.  
% FIXME 3: CONSIDER SHOWING DETAILED COMPARISON OF ESTIMATED VALUES TO ONES 
% EXPECTED FROM THE ACTUAL REC. TREES. THESE DON'T LOOK AS NICE, BUT THEY MIGHT 
% BE ABLE TO DEMONSTRATE OUR ABILITY TO DETECT DEPARTURE FROM EXPECTATION. SC:
% THIS MAY BE ABOUT CHANGING AVERAGE TO PRIOR EXPECTED NUMBER OF EVENTS.
% FIXME 4: WE SHOULD PROBABLY CONSIDER ALTERNATIVE WAYS OF USING THESE RESULTS TO 
% ARGUE THAT OUR DIRECTIONAL SIGNAL IS REAL. SC: I MIGHT CONSIDER SIMULATIONS AND
% ARTIFICIALLY DOUBLE OR TRIPLE THE RECOMBINATION EVENTS OF ONE OF TYPES.
We checked if numbers of recombinant edge types were inferred correctly. Because
of the two-stage Bayesian method we first sampled ten recombinant trees given
the mutation rate, recombination rate, and tract length (see section
\textbf{Results}), and then created ten independent simulated data sets with the
Jukes-Cantor model for DNA sequence evolution using each of the recombinant
trees. This consisted of 100 replicates of data sets.  Figure S\ref{fig:h1}
shows the observed count of each recombinant edge type against the expected
count; the estimated numbers appeared to be slightly over-estimated, but in
correlation with the true numbers.

@@

\subsection{}

@@

We tested if our measure of recombination intensity provided a reliable estimate
of the true intensity. For this purpose, we artificially annotaded our synthetic
blocks to genes according to the real annotations along the real bacterial
alignments. For each gene in each of the simulated 100 data sets, we plotted the
average estimated recombination intensity along that gene against the true
average intensity, i.e., the average number of different recombinant edge types
implied by the recombinant tree used in simulation (Figure S\ref{fig:ri1}).
Although the scatter plot did not strictly follow along the $X=Y$ line, there
was correlation of the estimated and true values (Pearson's correlation
coefficient of 0.63). Estimated values appeared to be driven somewhat by the
prior implied by the population parameter setting, explaining why lower values
were over-estimated and higher values were under-estimated. Nonetheless, the
correlation indicated that the measure of recombination intensity used in our
functional gene category served as an adequate indicator of actual intensity of
recombination events along genes.

@@

%=============================================================================
% Figure S: 
\begin{figure}
\subfloat[]{\includegraphics[scale=0.5]{figures/h1-R}}
\subfloat[]{\includegraphics[scale=0.5]{figures/h1-R-inset}}
\caption{\label{fig:h1}Numer of recombination events. The diagonal dashed
lines are with slope of 1 and intercept of 0. For each order
pairs of branches of a reference species tree the true number of recombination
events is computed as the mean of the 10 replicates of recombinant
trees from the simulation. With ten estimates of each replicate total
number of 100 estimats are shown as intervals of 90\% of quantile
range, or the lower value of the interval is the 5\% quantile, and
the upper one is the 95\% quantile. }
\end{figure}
\clearpage{}%
%=============================================================================

%=============================================================================
% Figure S:
\begin{figure}
\includegraphics{figures/ri1}
\caption{\label{fig:ri1}The surrogate measure of recombination intensity
of genes. Each point represents a pair of a true value of average number of
recombinant edges and its estimated value for a gene. The dashed line is with
slope of 1 and intercept of 0.}
\end{figure}
\clearpage{}%
%=============================================================================

%=============================================================================
% Figure S:
\begin{figure}
\subfloat[]{\includegraphics[scale=0.5]{figures/h2-R}}
\subfloat[]{\includegraphics[scale=0.5]{figures/h2-R-inset}}

\subfloat[]{\includegraphics[scale=0.5]{figures/h2-R-inset2}}

\caption{\label{fig:h2}Comparison of number of recombinations between real
and simulated data sets. The diagonal dashed lines are with
slope of 1 and intercept of 0. For each order pairs of branches of
a reference species tree the number of recombination events is estimated
from the real data set. Total number of 100 estimats from a simulated
data set are shown as intervals of 90\% of quantile range, or the
lower value of the interval is the 5\% quantile, and the upper one
is the 95\% quantile. }
\end{figure}
\clearpage{}%
%=============================================================================

\clearpage

@@

>>>>

\section*{Abstract}

Homologous recombination and horizontal gene transfer are investigated with
complete genomes of closely related bacterial species of \textit{Streptococcus
pyogenes} (SPY) and \textit{S.\ dysgalactiae} subspecies \textit{equisimilis}
(SDE).  The former is a pervasive human-specific pathogen, and the latter an
opportunistic human pathogen.  We aims at which parts of the genomic
organization and human pathogenic virulence factors of SDE could have originated
in those of SPY, or vice versa.  Our genome-wide survey of gene transfer
indicated that gene transfer via homologous recombination occurred more often
for the net direction from SPY to SDE.  We found evidence that virulence genes
transferred more often via non-recombining fashion without replacing parts of
destination genomes than did non-virulence genes.  A genome-wide resource of
recombining and non-recombining gene transfer for the study of the SPY and SDE
genomes is provided so that research community in the pathogenic SPY and SDE
bacteria could take advantage of it.

@@

Bacteria evolves new biological features through horizontally transferred genes
from other lineages and vertically transmitted mutations on genes within the
same lineage.  While mutations on a gene modifiy relatively small portions of the 
gene through substitutions, insertions, and deletions of deoxyribonucleic acids,
gene transfers can reshape the organization of a bacterial genome,
which could revamp biological functions of the species.  Recipient bacteria of
transferred genes can substitute transferred genes for homologous parts of their
genomes, or integrate transferred genes into their genomes without replacement.
These two processes are here in this manuscript discriminated by calling the
former \textit{recombining gene transfer}, and the latter \textit{horizontal
gene transfer} \citep{Ochman2001,Lawrence2009}.  Despite the two distinct
evolutionary processes in bacteria, analyses of complete genome sequences in
efforts of better understanding bacterial evolution have paid less attention to
a combined approach to illuminating disparity of the two gene transfer
processes. We attempted here to use and compare two approaches for
distinguishing recombining gene transfer process from horizontal, or
non-recombining, gene transfer, and tackled the two processes in association with
virulence genes.

@@

In short, the complete genomes of the two spcies of SPY and SDE presented us a
unique opportunity to study gene transfer over relatively recent time scales.
We took advantage of the model-based approach for inferring homologous
recombination, and the parsimony-based one for detecting gene duplication, loss,
and horizontal transfer. We compared both of the results in recombining gene
transfer events to show that the parimony-based method inferred recombining gene
transfer events in biological functional categories similar to those inferred
with the model-based approach. This was the first attempt as far as we know to
analyze SPY and SDE in genome-wide scale for detecting gene transfer events not
only in the core genome but also in the pan genome, which made the analysis
complete in terms of the amount of used genome data. Our analysis of core
genomes also showed for the first time that the net directionality of
recombining gene transfer from SPY to SDE was stronger than the reverse
direction.

@@

\texttt{Limitation of ClonalOrigin: inability of inferring recombinant edges}

\texttt{Need of a new model with asymmetry of gene transfer}

\texttt{Need of more SDE genomes}

@@

We present a short summary of the statistical method developed by
\citet{Didelot2007} and \citet{Didelot2010} for studying bacterial
recombination. A detailed account was found in the two papers and in the user
guide available at \url{http://code.google.com/p/clonalorigin}.  We used
\texttt{progressiveMauve} v2.3.1 \citep{Darling2004,Darling2010} with default
options to attempt to align the six genomes, which led to too small proportion
of core genomes. We had to remove the SEE sample from the genome alignment
although we kept it in the parsimony-based approach.  We then used
\texttt{stripSubsetLCBs}, a program of the \texttt{ClonalOrigin} package, and
identified 276 blocks of sequence alignments with length threshold of 1500 base
pairs.  We excluded two of these blocks in a subsequent analysis because they
contained long consecutive gaps.  We applied \texttt{ClonalFrame} v1.1 available
at \url{http://www.xavierdidelot.xtreemhost.com} to the remaining 274 blocks,
and estimated the species tree relating the five genomes. \texttt{ClonalFrame}
also estimates the ratio between recombination and mutation rates.
We used a single Markov chain Monte Carlo (MCMC) with 10000 burn-in generations
and 10000 sampling generations to obtain the tree estimate with branch lengths,
and additionally seven independent MCMC runs to validate adequate convergence of
the first run.

@@

@@

While \texttt{ClonalOrigin} could be used to understand recombining gene
transfers within the conserved core genome, we sought to understand the history
of the remainder of the genome where more complex evolutionary processes occur,
such as gene duplications, losses, and horizontal transfers.  In recent years,
several methods have been developed for inferring these evolutionary events in
gene families by reconciling gene trees with a species tree
\citep{David2011,Doyon2011,Tofigh2011}.  In this analysis, we complemented our
study of recombination by using \texttt{Mowgli} \citep{Doyon2011} to
characterize events that can change the copy number of a gene, such as
duplications, losses, and non-recombining horizontal gene transfers.  In
contrast to \texttt{ClonalOrigin}, \texttt{Mowgli} is parsimony-based for
inferring events and uses pre-determined gene clusters where genes are treated
as atomic units.  The advantage of using \texttt{Mowgli} was that it did not
assume the sequence data contains a single copy of each homologous segment, and
it attempted to explain different copy numbers by a series of gain and loss
events.  We obtained gene family clusters defined in \citet{Suzuki2011} using
the \texttt{OrthoMCL} method \citep{Li2003} over the full set of the six genomes
including \textit{S.\ equi} subsp.\textit{\ equi} strain 4047 (SEE).  For each
family cluster, the corresponding protein sequences were aligned using the
\texttt{MUSCLE} program \citep{Edgar2004a} and nucleotide alignments were
constructed by mapping nucleotide sequences onto each protein sequence
alignment. We then identified recombination within gene alignments using the
Single Breakpoint Recombination (\texttt{SBP}) method from the \texttt{HyPhy}
package \citep{KosakovskyPond2006}.  Using the Akaike Information Criterion
(AIC), \texttt{SBP} inferred that 47.5\% (1,094) of gene families contained at
least one topology-changing recombination. Each of these families was then split
into two gene fragments.  Gene trees were constructed from nucleotide alignments
for each gene fragment using the \texttt{RAxML} program \citep{Stamatakis2006},
and were rooted by minimizing inferred duplications and losses.  Each rooted
gene tree was then reconciled using the \texttt{Mowgli} program
\citep{Doyon2011}, which inferred the most parsimonious gene duplication, loss,
and transfer events for each family.  In the case of multiple maximum
parsimonious reconciliations, one was chosen uniformly at random.  Unless
specified otherwise, default setings were used for all the programs.  Since the
method of \texttt{Mowgli} did not directly model recombination events,
recombination-based transfers were inferred in a post-processing step, by
checking whether the transfer was coupled with a loss of a homologous gene in
the recipient branch (indicating gene replacement rather than gain).  A transfer
event was classified as non-recombining if the destination species of that
transfer had an extant descendant that contained copies of that genes which were
descendants of the transferred gene as well as copies which were
non-descendants.  We noted that this parsimonious approach was imperfect and
would tend to undercount events, especially recombining gene transfer events,
which were modeled using two events: a transfer and a loss.  Nevertheless, this
analysis allowed us to study non-recombining gene transfers, which were not
modeled by \texttt{ClonalOrigin}, and to validate some of the
\texttt{ClonalOrigin} results on recombining gene transfer.

@@

We performed a series of analyses to check for functional gene categories
enriched for certain types of gene transfer events. Functional categories were
assigned in the same manner as \citet{Suzuki2011}. In short, \texttt{blastP} 
was used to
compare Streptococcus genes to bacterial proteins from the Uniref90 database.
Matches with E-values less than $1.0\times10^{-5}$ were assumed to have the same
Gene Ontology (GO) classification as the target gene, as determined from the
uniProt GOA database.  We also classified a category of  \textit{virulent genes}
using the virulence profile described in \citet{Suzuki2011}. Among 207 virulent
genes at least one of the genes in the gene family was identified by that
profile.  Each gene covered by our analysis was associated with a recombination
intensity score from the \texttt{ClonalOrigin} analysis and scores from the
\texttt{Mowgli} analysis corresponding to number of duplications, losses,
horizontal gene transfers, and recombining gene transfer events.  Mann-Whitney
tests were used to identify GO categories whose genes showed significantly
elevated values in each score.  A 5\% False Discovery Rate (FDR) correction
\citep{Benjamini1995} was used to determine an appropriate significance
threshold for our p-values.

@@

We conducted experiments on simulation data to check the accuracy of estimates
obtained by \texttt{ClonalOrigin} in a scenario similar to one from the real 
data set.  We used a
simulation feature of \texttt{ClonalOrigin} to simulate the process of mutation
and recombination across the five species 
(see section \textbf{Supplementary Information} for detail).
While populaiton mutation and
recombination parameters were accurately estimated, the recombinant tract length
was over-estimated  (Table S\ref{tab:sim-three-population-parameters}).  This
was not surprising as recombinant tract lengths were known to be difficult to
estimate from relatively short alignment blocks \citep{Didelot2010}. 
We also evaluated the accuracy of the statistics estimated in the second stage
of \texttt{ClonalOrigin}. The number of recombinant edges for each pair of
recombinant tree branches was correlated with the true value in general while
there were slight over-estimation (Figure S\ref{fig:h1}).  However, this
comparison between inferrence from simulated data and that of real data set
suggested that the directional difference between SPY and SDE was due to neither
lack of power of \texttt{ClonalOrigin} nor hidden bias thereof.  Recombination
intensities that were measured on genes also showed rough correlation (Pearson's
correlation coefficient of 0.63) of the estimated and true values (Figure
S\ref{fig:ri1}).  Estimated values appeared to be driven somewhat by the prior
implied by the population parameter setting, explaining why lower values were
over-estimated and higher values were under-estimated.  Nonetheless, the
correlation indicated that the measure of recombination intensity used in our
functional gene category served as an adequate indicator of actual intensity of
recombination events along genes.

@@

\subsection{Local gene trees and recombination intensity}

Because the procedure of summarizing posterior samples of \textit{recombinant
tree}, which will be defined in a moment, 
was not detailed in the user guide at the time when we used
\texttt{ClonalOrigin}, we elected to use Figure S\ref{fig:clonalorigin} to
illustrate a recombinant tree in order to clarify summarization of
\textit{recombinant history} posterior samples.  A recombination history is
represented by a \textit{recombinant tree}, which is a species tree augmented
with \textit{recombinant edges}.  A recombinant edge connect two species tree
branches.  Each recombinant edge is defined by the time
and branch of origin and destination, and the genomic interval that it affects.
The branch of origin is referred to as the \textit{donor} and the destination
branch is referred to as the \textit{recipient}. The type of a recombinant edge
is determined by the donor and recipient species tree branches.  A recombinant
tree can be viewed as a restricted version of an ancestral recombination graph
(ARG), and as such it induces a local genealogy for each site in the genomic
block it describes.  \citet{Didelot2010} presented two procedures of summarizing
recombinant trees by viewing trees from two different angles: by counting types
of recombinant edges regardless of genomic positions affected by the edges, and
by measuring recombinant edge changes to the species tree in site-wise manner.
By similarly following their approaches we quantified additional aspects of
recombinant trees; we counted gene tree topologies for each site, and also
quantified recombination intensity along the genome.

The recombination intensity allowed us to find genomic regions enriched for
recombination events in general, or regions enriched for recombination events of
certain types (e.g., between SPY and SDE).  We noted that the recombination
intensity was not normalized to the interval of $[0,1]$. Rather than giving
probabilistic meanings to the recombination intensity we interpreted it as an
average number of recombinant edge types per site.  The measure appeared to act
as an effective indicator of recombination intensity (see subsection of
\textbf{Simulation Study}).  Our simulation study indicated that edges of
certain types were better indicators of recombination events than others.
For instance, recombinant edges which resulted in a recombinant tree with the
same topology as the species tree appeared to be more prone to spurious
sampling.  Such edges were ones where the donor was the same branch, sister
branch, or parent branch of the recipient branch.  Therefore, in some cases we
elected to remove such edges from consideration when measuring recombination
intensity.  
% The tree topologies changes would be correlated with recombination
% intensity.  The tree topologies would show more effectively regions where
% recombination occurs along the genome (see subsection of \textbf{UCSC Genome
% Browser}).

@@

% FIXME: Newick tree in text in a file.
% FIXME: Publish the result and the data set with the code.
We describe the detailed procedure of simulations for checking if the three
population parameters were accurately recovered from the first stage of
\texttt{ClonalOrigin} MCMC.  Ten recombinant trees were sampled given the three
parameters and the species tree  set to the values inferred from the five
bacterial genomes.  Each of the recombinant trees was used to simulate the
Jukes-Cantor model for DNA sequence evolution to create 274 blocks of
alignments.  The block lengths configuration was the same as that of the blocks
used in the analysis of the five genomes.  The mean of the ten estimates of the
recombinant tract length was significantly higher than the true value of (919 bp
versus 744 bp) whereas the values of the mutation and recombination rates were
accurately estimated (Table S\ref{tab:sim-three-population-parameters}).
Recombinant tract lengths were known to be difficult to estimate from relatively
short alignment blocks \citep{Didelot2010}.

@@

% FIXME 1: IG: RESULTS FROM THIS POINT ON REQUIRE SOME REFINEMENT. SC: Let's see.
% FIXME 2: ``TRUE" TO BE EXPECTED VALUE
% RATHER THAN AVERAGE OVER THE 10 REC.  TREES. SC: OKAY. LET'S FIND THE PRIOR
% EXPECTED NUMBER OF EVENTS. THIS MUST BE GIVEN SOMEWHERE}.  
% FIXME 3: CONSIDER SHOWING DETAILED COMPARISON OF ESTIMATED VALUES TO ONES 
% EXPECTED FROM THE ACTUAL REC. TREES. THESE DON'T LOOK AS NICE, BUT THEY MIGHT 
% BE ABLE TO DEMONSTRATE OUR ABILITY TO DETECT DEPARTURE FROM EXPECTATION. SC:
% THIS MAY BE ABOUT CHANGING AVERAGE TO PRIOR EXPECTED NUMBER OF EVENTS.
% FIXME 4: WE SHOULD PROBABLY CONSIDER ALTERNATIVE WAYS OF USING THESE RESULTS TO 
% ARGUE THAT OUR DIRECTIONAL SIGNAL IS REAL. SC: I MIGHT CONSIDER SIMULATIONS AND
% ARTIFICIALLY DOUBLE OR TRIPLE THE RECOMBINATION EVENTS OF ONE OF TYPES.
Because of the two-stage Bayesian method we first sampled ten recombinant trees
given the mutation rate, recombination rate, and tract length, and then created
ten independent simulated data sets with the Jukes-Cantor model for DNA sequence
evolution using each of the recombinant trees. This consisted of 100 replicates
of data sets.  Figure S\ref{fig:h1} shows the observed count of each recombinant
edge type against the expected count; the estimated numbers appeared to be
slightly over-estimated, but in correlation with the true numbers.

@@

@@

We tested if our measure of recombination intensity provided a reliable estimate
of the true intensity. For this purpose, we artificially annotated our synthetic
blocks to genes according to the real annotations along the real bacterial
alignments. For each gene in each of the simulated 100 data sets, we plotted the
average estimated recombination intensity along that gene against the true
average intensity, i.e., the average number of different recombinant edge types
implied by the recombinant tree used in simulation 
(Pearson's correlation coefficient of 0.63 and Figure S\ref{fig:ri1}).

@@

%=============================================================================
% Figure S12: 
\begin{figure}
\subfloat[]{\includegraphics[scale=0.5]{figures/h3-R-ratio-topology-changing}}
\subfloat[]{\includegraphics[scale=0.5]{figures/h3-R-ratio-not-topology-changing}}
\caption{\label{fig:h3}Numer of recombination events. The diagonal dashed
lines are with slope of 1 and intercept of 0. For each order
pairs of branches of a reference species tree the true number of recombination
events is computed as the mean of the 10 replicates of recombinant
trees from the simulation. With ten estimates of each replicate total
number of 100 estimats are shown as intervals of 90\% of quantile
range, or the lower value of the interval is the 5\% quantile, and
the upper one is the 95\% quantile. }
\end{figure}
\clearpage{}%
%=============================================================================

%=============================================================================
% Figure S12: 
\begin{figure}
\subfloat[]{\includegraphics[scale=0.5]{figures/h1-R}}
\subfloat[]{\includegraphics[scale=0.5]{figures/h1-R-inset}}
\caption{\label{fig:h1}Numer of recombination events. The diagonal dashed
lines are with slope of 1 and intercept of 0. For each order
pairs of branches of a reference species tree the true number of recombination
events is computed as the mean of the 10 replicates of recombinant
trees from the simulation. With ten estimates of each replicate total
number of 100 estimats are shown as intervals of 90\% of quantile
range, or the lower value of the interval is the 5\% quantile, and
the upper one is the 95\% quantile. }
\end{figure}
\clearpage{}%
%=============================================================================

%=============================================================================
% Figure S13:
\begin{figure}
\includegraphics{figures/ri1}
\caption{\label{fig:ri1}The surrogate measure of recombination intensity
of genes. Each point represents a pair of a true value of average number of
recombinant edges and its estimated value for a gene. The dashed line is with
slope of 1 and intercept of 0.}
\end{figure}
\clearpage{}%
%=============================================================================

%=============================================================================
% Figure S:
%\begin{figure}
%\subfloat[]{\includegraphics[scale=0.5]{figures/h2-R}}
%\subfloat[]{\includegraphics[scale=0.5]{figures/h2-R-inset}}

%\subfloat[]{\includegraphics[scale=0.5]{figures/h2-R-inset2}}

%\caption{\label{fig:h2}Comparison of number of recombinations between real
%and simulated data sets. The diagonal dashed lines are with
%slope of 1 and intercept of 0. For each order pairs of branches of
%a reference species tree the number of recombination events is estimated
%from the real data set. Total number of 100 estimats from a simulated
%data set are shown as intervals of 90\% of quantile range, or the
%lower value of the interval is the 5\% quantile, and the upper one
%is the 95\% quantile. }
%\end{figure}
%\clearpage{}%
%=============================================================================

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Should we keep this section?}
